Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-31 paper.tex
revised intro and abstract

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disks such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. % with, for example, radial variations in temperature. 
  However, the VSI requires a short thermal or cooling 
  timescale in sub-adabatically (stably) stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is thus needed to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of stably stratified accretion disks in which the disk  
  temperature $T$ is relaxed to its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find that 
  increasing $\beta$ even slightly away from    
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  is the radial temperature gradient ($r$
  being the cylindrical radius from the central star), and 
  $\gamma$ is the adiabatic index. 
  We apply our results  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. In the inner parts of the MMSN, e.g. at $\lesssim 5$AU, 
  the VSI will neccessarily have extremely short lengthscales $\ll 
  O(10^{-2}H)$, where $H$ is the disk scale height. In the outer disk, 
  we find the MMSN is unstable to the VSI from $\sim 5$AU to $\sim
  50$AU with a characteristic lengthscale $O(10^{-2}H)$ to
  $O(10^{-1}H)$ and a growth time of $\sim 30$ local orbital periods.  
  % ,and the dominant VSI mode in the MMSN is the `fundamental
  % corrugation mode' which causes the disk column to oscillate
  % vertically.
  We thus suggest the VSI is dynamically important, at least in the 
  outer parts of protoplanetary disks at tens of AU. 
\end{abstract}

@@

However, MHD turbulence may not be applicable to all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent  numerical simulations
suggest significant portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. Thus, alternative routes 
to angular momentum transport in PPDs remain of interest
\citep{armitage10}.

One possibility is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height $z$ from the disk midplane. This
dependence occurs when the equilibrium pressure depends on both the
temperature and density, making the disk baroclinic. It is therefore
generally expected that $\p_z\Omega\neq 0$ in astrophysical disks.

The VSI can be regarded as the disk analog of the
Goldreich-Schubert-Fricke instability 
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, can render
inertial waves unstable. \cite{barker15} gives a simple 
interpretation of the VSI: in an axisymmetric inviscid disk, the 
exchange of fluid parcels vertically, while conserving their specific angular
momentum, can result in a new configuration of lower energy if the
distribution of specific angular momentum varies with $z$. The VSI is
characterized by perturbations with large vertical length-scales
(relative to radial length-scales).

Early analysis of the VSI in radially and vertically local 
models of accretion disks \citep{urpin98,urpin03} showed that the VSI
requires a short thermal timescale so that perturbations evolve 
isothermally, unless the disk is neutrally stratified in the vertical
direction. This is because vertical motions associated with the VSI
are stabilized by buoyancy forces if the disk is sub-adiabatically
stratified (convectively stable). A sufficiently short thermal
timescale renders buoyancy ineffective.

The need for a short thermal timescale in stably stratified disks has
been demonstrated in numerical simulations of the VSI
\citep{nelson13,stoll14}. When the  thermal timescale $t_c$ is
parametrized in units of the orbital period  $P_\mathrm{orb}$,
\cite{nelson13} find that $t_c\lesssim  0.01P_\mathrm{orb}$ was needed
to observe the VSI. That is, the  disk temperature needs to remain
effectively unchanged. Radiation hydrodynamic simulations performed by
\cite{stoll14} also showed that the VSI is only sustained with the
inclusion of heating from stellar irradiation, which results in an
effective thermal timescale $t_c\ll P_\mathrm{orb}$ that maintains a
fixed temperature profile.

Given that the VSI is generally sensitive to disk thermodynamics,  
it is important to understand and quantify this dependence, for
example on disk parameters, in order to evaluate the 
relevance of the VSI to realistic disks, in which one may expect
non-zero vertical entropy gradients and thermal timescales. Should a
given disk become turbulent through the VSI, it must first pass
through the linear instability. Thus, understanding the thermodynamic
dependence of the  linear VSI is essential.

@@

This paper is organized as follows. In \S\ref{setup} we list the
governing equations and global equilibrium disk models under
consideration. We formulate the linear stability problem in
\S\ref{linear} and simplify it by making the radially local
approximation. We study the linear problem using two approaches. In
\S\ref{analytical} we present an analytical discussion of vertically
isothermal disks. Our treatment is grossly simplified owing to a
number of approximations made, but closely parallels previous work and 
allows us to highlight the essential effects of vertical shear. In
particular, we quantify a characteristic upper limit to the thermal
timescale for the VSI to operate. In \S\ref{numerical} we numerically
solve the linear problem in the radially local approximation, but
relax all other approximations made in our analytical approach.    
We find basic agreement with our analysis, including
the aforementioned characteristic thermal timescale. In
\S\ref{application} we apply our linear models to the Minimum Mass
Solar Nebula (MMSN) by modeling thermal relaxation
timescales based on realistic disk properties, and find that the VSI may
operate in the outer parts of the MMSN. We summarize in
\S\ref{summary} with a discussion of caveats and possible directions
for future work.

>>>>

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disks such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. % with, for example, radial variations in temperature. 
  However, the VSI requires a short thermal or cooling 
  timescale in sub-adabatically (stably) stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is thus needed to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of stably stratified accretion disks in which the disk  
  temperature $T$ is relaxed to its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find that 
  increasing $\beta$ even slightly away from    
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  is the radial temperature gradient, and 
  $\gamma$ is the adiabatic index. 
  We apply our results  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. In the inner parts of the MMSN, e.g. at $\lesssim 5$AU, 
  we find the VSI would only operate on extremely short lengthscales $\ll 
  O(10^{-2}H)$, where $H$ is the disk scale height. On longer
  lengthscales $O(10^{-2}H)$ to $O(10^{-1}H)$, we find the MMSN is
  unstable to the VSI from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.  
  % ,and the dominant VSI mode in the MMSN is the `fundamental
  % corrugation mode' which causes the disk column to oscillate
  % vertically.
  We thus suggest the VSI is at least dynamically important in the 
  outer parts of protoplanetary disks at tens of AU. 
\end{abstract}

@@

However, MHD turbulence may not be applicable to all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent MHD simulations
suggest significant portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. Thus, alternative mechanisms 
to angular momentum transport in PPDs, such as purely hydrodynamic
turbulence,  remain of interest \citep{armitage10}. In addition to
mass accretion, hydrodynamic turbulence in PPDs will also
impact dust evolution and planetesimal formation \citep{chiang10}.

One possible route to hydrodynamic turbulence is the vertical shear
instability (VSI). The VSI is an 
axisymmetric linear instability 
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height $z$ from the disk midplane. This
dependence occurs when the equilibrium pressure depends on both the
temperature and density, making the disk baroclinic. It is therefore
generally expected that $\p_z\Omega\neq 0$ in astrophysical disks.

The VSI can be regarded as the disk analog of the
Goldreich-Schubert-Fricke instability 
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
A height-dependence in $\Omega$, or vertical shear, can render
inertial waves unstable. \cite{barker15} gives a simple 
interpretation of the VSI: in an axisymmetric inviscid disk, the 
exchange of fluid parcels vertically, while conserving their specific angular
momentum, can result in a new configuration of lower energy if the
distribution of specific angular momentum varies with $z$. The VSI is
characterized by perturbations with large vertical length-scales 
relative to radial length-scales.

Early analysis of the VSI in radially and vertically local 
models of accretion disks \citep{urpin98,urpin03} showed that the VSI
requires a short thermal timescale so that perturbations evolve 
isothermally, unless the disk is neutrally stratified in the vertical
direction. This is because
vertical motions associated with the VSI 
are stabilized by buoyancy forces if the disk is sub-adiabatically
stratified (convectively stable). A sufficiently short thermal
timescale renders buoyancy ineffective.    
%efficient exchang of heat with surroundings

The need for a short thermal timescale for the VSI to operate in
stably stratified disks has been demonstrated in recent numerical simulations 
\citep{nelson13,stoll14}. When the  thermal timescale $t_c$ is
parametrized in units of the orbital period  $P_\mathrm{orb}$,
\cite{nelson13} find that $t_c\lesssim  0.01P_\mathrm{orb}$ was needed
to observe the VSI. That is, the  disk temperature needs to remain
effectively unchanged. Radiation hydrodynamic simulations performed by
\cite{stoll14} also showed that the VSI is only sustained with the
inclusion of heating from stellar irradiation, which results in an
effective thermal timescale $t_c\ll P_\mathrm{orb}$ that maintains a
fixed temperature profile.

These studies show that the VSI is generally sensitive to disk
thermodynamics. It is thus important to quantify this dependence, for
example on disk parameters, in order to evaluate the relevance of the
VSI to realistic disks, which one may expect to be stably
stratified and have non-zero thermal timescales. For example, we show in  
Fig. \ref{bcrit_mmsn2d} an estimate of the thermal timescales in a
typical protoplanetary disk model (see below). In order to determine
where in the disk the VSI may operate, and hence lead to hydrodynamic
turbulence, understanding the thermodynamic dependence of the linear VSI is essential. 
% since the transition
% Should a disk
% become turbulent through the VSI, it must first pass through the linear
% instability.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/bcrit_mmsn2d} 
  \caption{Estimate of the thermal timescale  in a fiducial
    protoplanetary disk model, the Minimum Mass Solar Nebula, for
    perturbations with radial lengthscales $\sim 10^{-2}H$, where $H$
    is the disk scale-height. The red line corresponds to a 
    characteristic thermal timescale, derived in this work, 
    below which the VSI is expected to operate effectively. 
    \label{bcrit_mmsn2d} 
  }
\end{figure}

@@

This paper is organized as follows. In \S\ref{setup} we list the
governing equations and global equilibrium disk models under
consideration. We formulate the linear stability problem in
\S\ref{linear} and simplify it by making the radially local
approximation. We study the linear problem using two approaches. In
\S\ref{analytical} we present an analytical discussion of vertically
isothermal disks. Our treatment is grossly simplified owing to a
number of approximations made, but closely parallels previous work and 
allows us to highlight the essential effects of vertical shear. In
particular, we quantify a characteristic upper limit to the thermal
timescale for the VSI. In \S\ref{numerical} we numerically
solve the linear problem in the radially local approximation, but
relax all other approximations made in our analytical approach.    
We find basic agreement with our analysis, including
the aforementioned characteristic thermal timescale. In
\S\ref{application} we apply our results to the Minimum Mass
Solar Nebula (MMSN) by modeling thermal 
timescales based on realistic disk properties, and find that the VSI
can operate effectively in the outer parts of the MMSN. We summarize in
\S\ref{summary} with a discussion of caveats and possible directions
for future work.

### 2015-03-31 setup.tex
major revision in process because of barker & latter
revised intro and abstract

For most of this paper we will take $\beta$ as a constant input
parameter, so that we can vary the thermodynamic response of the
disk from isothermal ($\beta\ll 1$) to adiabatic ($\beta \gg 1$) in a
controlled manner. However, for application to protoplanetary disks in 
\S\ref{application} we will model $\beta$ based on physical thermal
timescales, in which case $\beta$ is non-constant and is determined by
the equilibrium disk properties and the perturbation length-scales of
interest.

\subsection{Baroclinic disk equilibria}\label{eqm}
The equilibrium disk is steady and axisymmetric with density,
pressure and rotation profiles $\rho(r,z)$, $P(r,z)$ and
$\Omega(r,z)$, respectively. These functions satisfy  
\begin{align}
  0 &= \frac{1}{\rho}\frac{\p P}{\p z} + \frac{\p \Phi_*}{\p z},\label{vert_equilibrium}\\
  r\Omega^2&= \frac{1}{\rho}\frac{\p P}{\p r} + \frac{\p \Phi_*}{\p
    r}.\label{hori_equilibrium} 
\end{align} 
These equations allows us to choose any mid-plane density 
distribution $\rho_0(r)\equiv \rho(r,0)$. In this work we assume a
power-law profile, 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index.

@@

For discussion purposes it is useful to define the vertical buoyancy
frequency $N_z$ via
\begin{align}
  N_z^2 \equiv - c_p^{-1}g_z \frac{\p S}{\p z} = c_s^2 \left(\frac{\gamma -
      \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,   \label{nzsq_def}
\end{align}
where $c_p$ is the heat capacity at constant pressure, $g_z\equiv
\rho^{-1}\p_zP $ and $S\equiv c_p\ln{P^{1/\gamma}/\rho}$ is the
specific entropy. The second equality applies to the disk equilibria
described above. 
We are interested in
neutrally or sub-adiabatically stratified 
disks with $\Gamma\leq \gamma$ so that $N_z^2\geq0$.

@@

\subsection{Radially-local approximation}  
We consider perturbations with small radial length-scales 
so that we can ignore the radial variations in the background
structure and radial gradients. We thus evaluate all coefficients in
Eq. \ref{full_mass1}---\ref{full_energy1} at the fiducial radius
$r=r_0$. However, the full vertical dependence of the coefficients are
retained. Then we can assume perturbations have spatial dependence in
the form 
\begin{align}
  \delta \rho(r,z) \to \delta\rho(z)\exp{\ii k_r r}.
\end{align}
We take $k_r>0$ without loss of generality, and further assume $k_rr\gg1$ so that curvature terms can also be 
ignored.

>>>>

For most of this paper we will take $\beta$ as a constant input
parameter, so that we can vary the thermodynamic response of the
disk from isothermal ($\beta\ll 1$) to adiabatic ($\beta \gg 1$) in a
controlled manner. However, for application to protoplanetary disks in 
\S\ref{application} we will model $\beta$ based on physical thermal
timescales, in which case $\beta$ is non-constant and is determined by
the equilibrium disk properties and the perturbation length-scales of
interest (e.g. Fig. \ref{bcrit_mmsn2d}).

\subsection{Baroclinic disk equilibria}\label{eqm}
The equilibrium disk is steady and axisymmetric with density,
pressure and rotation profiles $\rho(r,z)$, $P(r,z)$ and
$\Omega(r,z)$, respectively. These functions satisfy  
\begin{align}
  0 &= \frac{1}{\rho}\frac{\p P}{\p z} + \frac{\p \Phi_*}{\p z},\label{vert_equilibrium}\\
  r\Omega^2&= \frac{1}{\rho}\frac{\p P}{\p r} + \frac{\p \Phi_*}{\p
    r}.\label{hori_equilibrium} 
\end{align} 
These equations allows us to choose any midplane density 
distribution $\rho_0(r)\equiv \rho(r,0)$. In this work we assume a
power-law profile, 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index.

@@

For discussion purposes it is useful to define the vertical buoyancy
frequency $N_z$ via
\begin{align}
  N_z^2 \equiv - c_p^{-1}g_z \frac{\p S}{\p z} = c_s^2 \left(\frac{\gamma -
      \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,   \label{nzsq_def}
\end{align}
where $c_p$ is the heat capacity at constant pressure, $g_z\equiv
\rho^{-1}\p_zP $ and $S\equiv c_p\ln{P^{1/\gamma}/\rho}$ is the
specific entropy. The second equality applies to the disk equilibria
described above. We are interested in sub-adiabatically stratified 
disks with $\Gamma < \gamma$ so that $N_z^2\geq0$. % , and the disk is
% convectively stable in the vertical direction.

@@

% This approach has been adopted
% in recent linear calculations of the VSI
% \cite{nelson13,mcnally14,barker15} because unstable modes have short
% radial lengthscales.

\subsection{Radially-local approximation}  
We consider perturbations with small radial length-scales 
so that we can ignore the radial variations in the background
structure and radial gradients. \cite{barker15} have validated this
approximation by comparison with the linear VSI in global cylindrical
disk geometry. We thus evaluate all coefficients in
Eq. \ref{full_mass1}---\ref{full_energy1} at the fiducial radius
$r=r_0$. However, the full vertical dependence of the coefficients are
retained. Then we can assume perturbations have spatial dependence in
the form 
\begin{align}
  \delta \rho(r,z) \to \delta\rho(z)\exp{\ii k_r r}.
\end{align}
We take $k_r>0$ without loss of generality, and further assume $k_rr\gg1$ so that curvature terms can also be 
ignored.
